Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-07-31 joint_inf.tex
Cleaning out addressed comments.
Getting to a point of confusion.

Assume $\fullAlignment$ is generated from the InvFels topology $\tau^*$ and with true generating branch fidelities $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ (Fig.~\ref{fig:farris-fels-top}a).
Let $\boldsymbol\xi = [\xi_j]_{j=1}^\nSiteSplits$ be the vector of most likely ancestral state splits.
Use $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case
\begin{equation}
\label{eq:exp-log-lik}
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\end{equation}
We give $\ell$ explicitly as \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.
For a fixed $\tau$, let $\hat{t}_\nCols$ maximize the left-hand side of \eqref{eq:exp-log-lik} and $\hat{t}$ maximize the right-hand side.
We show in the Appendix that $\hat{t}_\nCols \rightarrow \hat{t}$, allowing us to focus on only the right-hand side above.

@@

As data are generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same optimization code.
For joint inference, the estimates for $\hat{p}_w$ when $p_{x^*}$ and $p_{y^*}$ are both small look reasonable, with $\hat{p}_w$ increasing as $p_{y^*}$ increases, though Fig.~\ref{fig:empirical-bias} shows there is a systematic negative bias in this procedure even when the true branches are short.
To understand the quality of each fit, we report the range of the bias $\hat{p}_w-p_{y^*}$ where $p_{x^*}, p_{y^*} \le .1$.
%EM I'd suggest bringing this figure into the main text. It'd fit.
%das: :check:
There are discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}) due to the choice of which ancestral state splits are maximal.
Because of these discontinuities, an overall estimate of the bias is misleading, and these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
For joint inference, the errors given $p_{x^*}, p_{y^*} \le .1$ range from $[-2\times 10^{-2}, 3\times 10^{-3}]$ and for inference on the integrated likelihood are lower than machine tolerance showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.

\begin{figure}
\centering
\includegraphics[width=.95\textwidth]{empirical-bias}
\caption{
Bias in branch length estimation.
In regions where joint optimization should perform well with $p_{x^*}, p_{y^*} \le .1$, there is systematic bias toward shorter branch lengths.
}
\label{fig:empirical-bias}
\end{figure}

@@

We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of topological inconsistency: the joint inference procedure estimates zero-length branches given the correct topology.
%EM Suggest "Considering this zero length branch as a multifurcating topology..."
%EM Oops, somehow I misunderstood this the first time around. You're saying that the wrong topology, with internal branch shrunk to zero, will have the same likelihood. And that we can't exclude that it could have a greater likelihood with internal branch nonzero. If you want to make those points (or something about multifurcating) they need to be more explicit. Right now it reads like we found cases where the wrong topology has larger likelihood.
%das: maybe this is better?
%EM Seems like clarification might be justified in other places in the MS as well.
%das: also hedged a bit in the remainder of the MS.
Considering a tree with a zero-length branch as a multifurcating topology, the incorrect topology attains the same likelihood as the topology that generated the data by fixing this branch to have zero length.
Since the parameters with the highest likelihood given the generating topology include a zero-length branch, we cannot exclude the possibility that the incorrect topology with this branch having nonzero length is more likely to be observed, though we have not found regions where this is the case.
The regions of inconsistency we found arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
We see that this inconsistency occurs even if some branches are short.
This expands on the empirical findings of poor estimation given long branches in \citet{Sagulenko2018-xl} (their Figures~2 and~3).
%EM I suggest flipping this around and making it stronger, so that we lead with and emphasize the differences.
%EM I've made this a little spicier: happy with the first part?
%das: I like it
However, the problems are not just for long branches as \citet{Sagulenko2018-xl} imply: even when all branches are short there is a consistent bias, and that the bias is on the same order as the magnitude of the parameters (Fig.~\ref{fig:empirical-bias})

>>>>

Assume $\fullAlignment$ is generated from the InvFels topology $\tau^*$ and with true generating branch fidelities $t^*=\{x^*, y^*, x^*, y^*, y^*\}$ (Fig.~\ref{fig:farris-fels-top}a).
Let $\boldsymbol\xi = [\xi_j]_{j=1}^\nSiteSplits$ be the vector of most likely ancestral state splits.
Use $\ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi)$ to denote the expected per-site log-likelihood, which can be thought of as the infinite-length sequence case because
\begin{equation}
\label{eq:exp-log-lik}
\frac{1}{n}\log L_\nCols'(\tau, t; \fullAlignment) \rightarrow \ell_{\tau^*,t^*}(\tau, t; \boldsymbol\xi).
\end{equation}
We give $\ell$ explicitly as \eqref{eq:site_pattern_profile_likelihood_mean} in the Appendix.
For a fixed $\tau$, let $\hat{t}_\nCols$ maximize the left-hand side of \eqref{eq:exp-log-lik} and $\hat{t}$ maximize the right-hand side.
We show in the Appendix that $\hat{t}_\nCols \rightarrow \hat{t}$, allowing us to focus on only the right-hand side above.

@@

As data are generated with parameters $\{x^*, y^*, x^*, y^*, y^*\}$, the true value for $w$ is $y^*$.
Marginal inference performs as expected, where $\hat{w}$ is equal to $y^*$ regardless of the value of $x^*$ (Fig.~\ref{fig:bl-general-marginal}) when optimizing \eqref{eq:marginal_likelihood} using the same optimization code.
For joint inference, the estimates for $\hat{p}_w$ when $p_{x^*}$ and $p_{y^*}$ are both small look reasonable, with $\hat{p}_w$ increasing as $p_{y^*}$ increases, though Fig.~\ref{fig:empirical-bias} shows there is a systematic negative bias in this procedure even when the true branches are short.
To understand the quality of each fit, we report the range of the bias $\hat{p}_w-p_{y^*}$ where $p_{x^*}, p_{y^*} \le .1$.
There are discontinuities in the fit (Fig.~\ref{fig:bl-general-inconsistency}) due to the choice of which ancestral state splits are maximal.
Because of these discontinuities, an overall estimate of the bias is misleading, and these short-branch cases should be the best settings for joint optimization \citep{Sagulenko2018-xl}.
For joint inference, the errors given $p_{x^*}, p_{y^*} \le .1$ range from $[-2\times 10^{-2}, 3\times 10^{-3}]$ and for inference on the integrated likelihood are lower than machine tolerance showing that, even in cases where joint inference is supposed to do well, it still fails to achieve a low error from truth.

\begin{figure}
\centering
\includegraphics[width=.95\textwidth]{empirical-bias}
\caption{
Bias in branch length estimation.
Even in regions with short branch length ($p_{x^*}, p_{y^*} \le .1$) where joint optimization should perform well, there is systematic bias toward shorter branch lengths.
}
\label{fig:empirical-bias}
\end{figure}

@@

We have shown that jointly inferring ancestral states and phylogenetic parameters \citep{Sagulenko2018-xl} is not consistent in general.
Specifically, in the case of four-taxon trees with infinite data, we have obtained nontrivial regions of generating parameters that result in a type of topological inconsistency: the joint inference procedure estimates zero-length branches given the correct topology.
Considering a tree with a zero-length branch as a multifurcating topology, the incorrect topology attains the same likelihood as the topology that generated the data by fixing this branch to have zero length.
Since the parameters with the highest likelihood given the generating topology include a zero-length branch, we cannot exclude the possibility that the incorrect topology with this branch having nonzero length is more likely to be observed, though we have not found regions where this is the case.
The regions of inconsistency we found arise when the top two branches of the generating trees are ``long,'' that is, when the top branch fidelities tend to be small, and when the lower branches are ``short,'' i.e., have large fidelities.
We see that this inconsistency occurs even if some branches are short.
This expands on the empirical findings of poor estimation given long branches in \citet{Sagulenko2018-xl} (their Figures~2 and~3).
However, the problems are not just for long branches as \citet{Sagulenko2018-xl} imply: even when all branches are short there is a consistent bias, and that the bias is on the same order as the magnitude of the parameters (Fig.~\ref{fig:empirical-bias}).
